---
title: "Data Visualization 2"
description: "Analysis of Chocolate Data from TidyTuesday"
author: William Walz
date: September 22, 2025
format: html
execute: 
  warning: false
  message: false
---

This analysis uses the TidyTuesday chocolate ratings dataset which contains reviews many different chocolate bars. I focus on the countries of cocoa bean origin to see which regions appear most often in the data. The plot below shows the ten most common bean origins.

```{r}
library(tidyverse)

chocolate <- read_csv("chocolate.csv")

top_bean_countries <- chocolate |>
  filter(!is.na(country_of_bean_origin)) |>
  count(country_of_bean_origin, sort = TRUE) |>
  head(10)

top_bean_countries
```

```{r}
ggplot(top_bean_countries, aes(x = reorder(country_of_bean_origin, n), y = n)) +
  geom_col(fill = "chocolate4") +
  coord_flip() +
  labs(
    title = "Top 10 Countries of Cocoa Bean Origin",
    x = "Country of Bean Origin",
    y = "Number of Reviews"
  ) +
  theme_minimal()
```


The TidyTuesday data can be found at: 
https://github.com/rfordatascience/tidytuesday/blob/main/data/2022/2022-01-18/chocolate.csv

This data comes from Flavors of Cacao by way of Georgios and Kelsey. (original link):
https://flavorsofcacao.com/chocolate_database.html







