---
title: "Data Visualization 1"
description: "Analysis of Artists Data from TidyTuesday"
author: William Walz
date: September 22, 2025
format: html
execute: 
  warning: false
  message: false
---

This dataset contains information on artists, including their names, genders, birth years, and places of origin. Here, I created a visualization the top 5 artist birth years to find the years that produced the most artists. The bar chart at the bottom displays the five most common years of birth in the dataset.\

```{r}

library(tidyverse)

artists <- read_csv("artists.csv")

top_birth_years <- artists |>
  filter(!is.na(yearOfBirth)) |>
  group_by(yearOfBirth) |>
  summarize(n = n()) |>
  arrange(desc(n)) |>
  head(5)

top_birth_years

```

```{r}

ggplot(top_birth_years, aes(x = reorder(as.factor(yearOfBirth), n), y = n)) +
  geom_col(fill = "steelBlue") +
  labs(
    title = "Top 5 Years of Artist Births",
    x = "Year of Birth",
    y = "Number of Artists"
  ) +
  theme_minimal()

```

The TidyTuesday data can be found at: <https://github.com/rfordatascience/tidytuesday/blob/main/data/2021/2021-01-12/artists.csv>

This data comes from the Tate Art Museum (original GitHub link): <https://github.com/tategallery/collection>
