---
title: "Data Visualization 1"
description: "Analysis of Artists Data from TidyTuesday"
author: William Walz
date: September 22, 2025
format: html
execute: 
  warning: false
  message: false
---

This dataset contains information on artists, including their names, genders, birth years, and places of origin. Here, I created a visualization of the top 5 artist birth years to find the years that produced the most artists. The bar chart at the bottom displays the five most common years of birth in the dataset.

Accessibility Note: All data visualizations in this report include alternative text (alt text) descriptions to ensure the findings are accessible to users with screen readers.\

```{r}

library(tidyverse)

artists <- read_csv("artists.csv")

top_birth_years <- artists |>
  filter(!is.na(yearOfBirth)) |>
  group_by(yearOfBirth) |>
  summarize(n = n()) |>
  arrange(desc(n)) |>
  head(5)

top_birth_years

```

```{r}
#| fig-alt: "A vertical bar chart displaying the five most common birth years for artists in the dataset. The tallest bars indicate the specific years when the highest number of artists in this collection were born."

ggplot(top_birth_years, aes(x = reorder(as.factor(yearOfBirth), n), y = n)) +
  geom_col(fill = "steelBlue") +
  labs(
    title = "Top 5 Years of Artist Births",
    x = "Year of Birth",
    y = "Number of Artists"
  ) +
  theme_minimal()
```

**Data Sources:**

-   The TidyTuesday data can be found at: [TidyTuesday 2021-01-12](https://github.com/rfordatascience/tidytuesday/tree/main/data/2021/2021-01-12)
-   Original data source: [Tate Art Museum Collection](https://github.com/tategallery/collection)
