---
title: "Project 5: Analzying US Traffic Stops"
description: "SQL-based analysis of stop demographics, violations, and search outcomes"
author: William Walz
date: December 6, 2025
format: 
  html:
    code-fold: true
    code-summary: "Show code"
execute: 
  warning: false
  message: false
---

Police traffic stops generate large amounts of administrative data: when and where each stop occurred, the driver's reported race, reason for the stop, and whether any searches or contraband findings occurred. The Stanford Open Policing Project (Pierson et al., 2020) compiles these records from a multitude of agencies across the United States into one database. For this project, I use SQL to explore three distinct aspects of traffic stop patterns. 

Specifically, I analyze [1] how the racial composition of traffic stops in New York City changed between 2010–2018, [2] the most frequently recorded traffic violations in Houston, Texas, and [3] the relationship between search rates and "hit rates" (contraband found by police) by race in Florida. All of the data wrangling is conducted in SQL. R is used only to visualize or format the summarized results. 

Each section below uses a different table, includes SQL queries only for wrangling, and presents a table or labeled figure, followed by a narrative explanation of what the output shows.

### Racial Distribution of Traffic Stops Over Time: New York State

To begin the analysis, I examine how the racial composition of recorded traffic stops in New York State has changed over time. The goal here is descriptive rather than causal. I summarize the data as it appears in the tables without making claims about why these patterns occur or what underlying mechanisms may explain them. 

The SQL query below groups stops by race and year from 2010-2018 and counts the number of stops in each category. Because the total number of stops varies much across years, I convert these counts into percentages so that each group’s share of all stops can be compared consistently over time. 

It is important to note that these percentages reflect the distribution within the stop dataset itself, not the demographic composition of New York State’s driving population, nor the population as a whole. Therefore, the results show how the dataset is structured rather than making any statement about disparities or policing practices.

```{r}
library(DBI)
library(RMariaDB)
library(dbplyr)

con_traffic <- DBI::dbConnect(
  RMariaDB::MariaDB(),
  dbname = "traffic",
  host = Sys.getenv("TRAFFIC_HOST"),
  user = Sys.getenv("TRAFFIC_USER"),
  password = Sys.getenv("TRAFFIC_PWD")
)
```


```{sql, connection=con_traffic, output.var="ny_stops"}
SELECT
  subject_race AS race,
  COUNT(*) AS n_stops,
  YEAR(date) AS year
FROM ny_statewide_2020_04_01
WHERE subject_race IS NOT NULL
GROUP BY race, year
HAVING year BETWEEN 2010 AND 2018;
```

*Figure 1. Traffic Stops in New York City by Race and Year as %*

```{r}
library(ggplot2)
library(bit64)
library(dplyr)
library(scales)

ny_stops_clean <- ny_stops |>
  mutate(across(where(bit64::is.integer64), as.integer))

ny_stops_pct <- ny_stops_clean |>
  group_by(year) |>
  mutate(
    total_stops_year = sum(n_stops),
    pct = n_stops / total_stops_year * 100
  ) |>
  ungroup()

ggplot(ny_stops_pct, aes(x = year, y = pct, color = race)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  labs(
    title = "Percent of Traffic Stops in New York State by Race and Year",
    x = "Year",
    y = "Percent of Stops",
    color = "Race"
  ) +
  theme_minimal(base_size = 14)
```

*What does this mean?*

The figure shows how the racial composition of recorded traffic stops in New York State changed between 2010 and 2018. Across all of the years, the majority of stops in this dataset involved individuals classified as white, with their share gradually declining from roughly 78% in 2010 to around 70% by 2018. Other groups make up much smaller proportions of the dataset, though their relative shares show slight increases over time. For example, stops recorded as Black rise from about 10% to approximately 15%, and Hispanic stops increase from roughly 6% to about 9% by 2018. 

It is important to emphasize that these percentages describe the distribution of stops within the SOPP dataset, not the underlying driving population or the general population of New York State. Without denominators (such as the racial composition of licensed drivers or traffic volume data) we cannot draw conclusions about disparities or exposure to police stops. The figure should therefore be interpreted strictly as a summary of how the stop records themselves are distributed across racial categories over time, rather than as evidence of the causes or consequences of those patterns.

### Most Common Types of Traffic Stops: Long Beach and San Bernardino, California

For the second part of the analysis, I compare how different agencies categorize traffic stops. Both Long Beach and San Bernardino report stop types, but neither provides detailed violation codes of what each specifically means. Instead, each uses broad categories such as vehicular or pedestrian, or other (bikes, scooters, etc.). The following SQL queries count how many times each stop type appears in each agency’s data.

```{sql, connection=con_traffic, output.var="lb_types"}
SELECT
type AS stop_type,
COUNT(*) AS type_count,
'Long Beach' AS city
FROM ca_long_beach_2020_04_01
GROUP BY stop_type;
```

```{sql, connection=con_traffic, output.var="sb_types"}
SELECT
type AS stop_type,
COUNT(*) AS type_count,
'San Bernardino' AS city
FROM ca_san_bernardino_2020_04_01
GROUP BY stop_type;
```


*Visualization: Comparing Stop Type Distributions Across Two Agencies*

```{r}
library(ggplot2)
library(dplyr)
library(scales)

# Combination of Long Beach & San Bernardino results

combined_types <- bind_rows(lb_types, sb_types) |>
  mutate(
    stop_type = ifelse(is.na(stop_type) | stop_type == "", "NA", stop_type)
    ) |>
  group_by(city) |>
  mutate(stop_type = reorder(stop_type, type_count)) |>
  ungroup()

ggplot(combined_types, aes(x = stop_type, y = type_count)) +
  geom_col(fill = "#4C72B0") +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~ city, nrow = 1, scales = "free_x") +
  labs(
    title = "Distribution of Stop Types Across Two California Agencies",
    subtitle = "Long Beach vs. San Bernardino",
    x = "Stop Type",
    y = "Number of Stops"
    ) +
  theme_minimal(base_size = 14)
```


*What does this mean?*



then for the interpretation after long beach is high level pedestriaan. na could be bikes, scooters, 

### Search Rates and Frisk Rates by Race: Florida

In the final part of this project, I analyze how often police searches and frisks occur during traffic stops in Florida, and how those decisions differ by race. The dataset includes search_conducted (whether any search occurred) and frisk_performed (whether a pat-down occurred). 

Before computing rates, I combined ambiguous racial categories such as NA, "unknown", and "other" into a single group labeled "unknown/other". These entries are recorded inconsistently and do not correspond to a clearly defined racial group. In the raw data, for example, the NA category had an unusually high frisk rate, which almost certainly reflects missing or miscoded values rather than a meaningful disparity. Collapsing these ambiguous categories prevents over-interpreting noisy or incomplete data.

Search + Frisk Rates (SQL)

```{sql connection=con_traffic, output.var="fl_search"}
SELECT
CASE
WHEN subject_race IS NULL OR subject_race IN ('other', 'unknown', '') THEN 'unknown/other'
ELSE subject_race
END AS race,
COUNT(*) AS total_stops,
SUM(CASE WHEN search_conducted = 1 THEN 1 ELSE 0 END) AS num_searches,
SUM(CASE WHEN frisk_performed = 1 THEN 1 ELSE 0 END) AS num_frisk,
1.0 * SUM(CASE WHEN search_conducted = 1 THEN 1 ELSE 0 END) / COUNT(*) AS search_rate,
1.0 * SUM(CASE WHEN frisk_performed = 1 THEN 1 ELSE 0 END) / COUNT(*) AS frisk_rate
FROM fl_statewide_2020_04_01
GROUP BY race;
```

*Visualization: Search & Frisk Rates*

```{r}
library(dplyr)
library(ggplot2)
library(scales)
library(tidyr)

fl_search_long <- fl_search |>
  select(race, search_rate, frisk_rate) |>
  mutate(race = factor(race)) |>
  pivot_longer(
    cols = c(search_rate, frisk_rate),
    names_to = "metric",
    values_to = "rate"
  ) |>
  mutate(
    metric = recode(
      metric,
      "search_rate" = "Search rate",
      "frisk_rate"  = "Frisk rate"
    )
  )

ggplot(fl_search_long,
       aes(x = race, y = rate, fill = race)) +
  geom_col() +
  facet_wrap(~ metric, ncol = 2, scales = "free_x") +
  coord_flip() +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  labs(
    title = "Search and Frisk Rates by Race in Florida",
    x = "Race",
    y = "Share of Stops Involving a Search or Frisk"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    strip.text = element_text(size = 12, face = "bold"),
    panel.spacing.x = unit(3, "lines")
  )
```



*What does this mean?*



***Conclusion***






**Sources**

Stanford Open Policing Project (SOPP) Data. 
Traffic stop data compiled by the Stanford Open Policing Project and accessed through the Pomona College SQL server. Original dataset and documentation available at: [https://openpolicing.stanford.edu](https://openpolicing.stanford.edu). Pierson et al. (2020).

New York City Demographic Data. 
U.S. Census Bureau. New York, New York – Profile (Profile ID 16000US3651000). 
Demographics can be found at: [https://censusreporter.org/profiles/16000US3651000-new-york-ny/](https://censusreporter.org/profiles/16000US3651000-new-york-ny/)

(Note: CensusReporter provides a visualization layer, but all values originate from the U.S. Census Bureau.)

***References***



